---
title: "MotorTrendAnalysis"
author: "Bently"
date: "1/28/2021"
output:
  html_document: default
  pdf_document: default
---
    
    *Created with rMarkdown*
    
#### Overall Summary
This report will analyze relationships between different Transmission compared to Miles/Gallon (MPG). 
The goal here is to show which transmission produces the best or a higher MPG. 

The [mtcars](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars) dataset was used for this analysis. 

When we do a t-test of automatic and manual transmission, manual transmission vehicles could have a 7.245 greater MPG 
over automatic transmission. In addition fitting multiple linear regressions, analysis of manual transmission contributed less
significantly MPG, a slight improvement of 1.81 MPG.  The other correlated variables, weight, horsepower, and no. of cylinders 
contributed more to overall MPG of vehicles.

#### Load Data
First we will need to load the dataset and set our variables to factors.
```{r message=FALSE}
library(ggplot2)
data(mtcars)
    mtcars$cyl <- as.factor(mtcars$cyl)
    mtcars$vs <- as.factor(mtcars$vs)
    mtcars$am <- factor(mtcars$am)
    mtcars$gear <- factor(mtcars$gear)
    mtcars$carb <- factor(mtcars$carb)
attach(mtcars)
cat("Dimension", dim(mtcars))
head(mtcars, n=3)

```

#### Exploratory Analysis
**See plot below**
A Box graph compares Automatic and Manual transmission MPG. The graph can show there is a significant increase in MPG when 
for vehicles with a manual transmission.

##### Statistical Inference
Here is our T-Test Transmission & MPG
```{r}
tResults <- t.test(mpg ~ am)

cat("T-Test P-value is ", tResults$p.value)

boxplot(mpg ~ am, 
        xlab="Transmission Type (0 = Automatic, 1 = Manual)", 
        ylab="MPG",
        main="MPG by Transmission Type")
```
The T-Test rejects the null hypothesis that the difference between transmission types is 0.  
```{r}
tResults$estimate
```
2 transmissions difference estimate is 7.24494 MPG in favor of manual transmissions.

##### Regression Analysis
Fit the full model 
```{r results='hide'}
lm.Regression <- lm(mpg ~ ., data = mtcars)
summary(lm.Regression)  # results hidden
summary(lm.Regression)$coeff  # results hidden
```
No coefficient had a p-value less than 0.05 we cannot deduce which variables are more significant. 

Now lets try Stepwise Regression to determine which variables are most significant
```{r results='hide'}
Stepwise <- step(lm.Regression)
summary(Stepwise) # results hidden
summary(Stepwise)$coeff # results hidden
```

Now we have 4 variables (cylinders, horsepower, weight, transmission). The R-squared value of 0.8659. 
- The p-values are less than 0.05. 
- The coefficients conclude that increasing the number of cylinders from 4 to 6 with decrease the MPG by 3.03.  
- Further increasing the cylinders to 8 with decrease the MPG by 2.16.  
- Increasing the horsepower is decreases MPG 3.21 for every 100 horsepower.  
- Weight decreases the MPG by 2.5 for each 1000 lbs increase. 
- A Manual transmission improves the MPG by 1.81.

#### Residuals & Diagnostics
Residual Plot
**See plot below**
    
    Us this plot to deduce:
    
1. Randomness of the Residuals vs. Fitted plot supports the assumption of
independence
2. The points of the Normal Q-Q plot following closely to the line conclude that
the distribution of residuals is normal
3. The Scale-Location plot random distribution confirms the constant variance 
assumption
4. Since all points are within the 0.05 lines, the Residuals vs. Leverage 
concludes that there are no outliers
```{r}
sum((abs(dfbetas(Stepwise)))>1)

par(mfrow = c(2, 2))
plot(Stepwise)
```

#### Conclusion
There is a difference in MPG based on transmission type. A manual transmission
will have a slight MPG boost. However, it seems that weight, horsepower, & 
number of cylinders are more statistically significant when determining MPG.
